Estimating Comparative Advantage in Improved Seed Adoption: Hybrid Maize in Ethiopia

### 2021-12-14 paper.tex
Updates from Overleaf

%%% Why it is important
Food security is a major challenge in many East African countries-- Ethiopia is no exception, having historically struggled to provide an adequate and reliable food supply \citep{Ramakrishna2002-hv, Jaleta2018-oj}. Since the drought of 1984, Ethiopia has become increasingly reliant on maize, as of today, it constitutes the most widespread crop in the country.\footnote{FAO stats data shows the total area cultivated with Maize in 2019 is 25\% larger than the one cultivated with the next most widespread crop (sorghum), and 27\% and 139\% larger than the next two following crops, wheat and Barley, respectively.} As Figure \ref{fig:maize_yields} shows, there is an upward trend in the total area cultivated by maize, with average yields that have almost double during the period.

%%%%% The puzzle 
Despite numerous technological breakthroughs in maize germplasm with increased yields and drought tolerance, adoption of improved seed has remained a challenge, with only 13\% of maize farmers consistently using improved seeds varieties, and the lion share of farmers (62\%) not using improved seeds at all during the period where data is available. Moreover, there is a large percentage of farmers who adopt new varieties inconsistency limiting their opportunities to learn about their benefits, see Table \ref{tbl:Trajectories}.

@@

The advantage of our estimation strategy is that we do not rely on identification from a valid instrument, but modelling the heterogeneity directly through a projection of their returns to adoption on their observed "trajectory" of adoption. This allows us to directly capture two effects: the heterogeneous effects of the so-called "comparative advantage" of adoption, and, under the assumption of linearity in comparative advantage, a description of the overall economy's sorting of farmers based on their comparative advantage. The first is important because it is able to disentangle the effect of a general household's agricultural ability (which can be thought of as a fixed effect) and their adoption-specific potential. The sorting parameter can tell us whether the economy sorts such that those that would benefit the most adoption actually adopt, or whether there are barriers that stop that sorting from happening. See \cite{Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc} for more information on the assumptions involved for the Group Random Coefficients estimator. In our particular case, we contribute to the li

@@

\begin{align}
y_{it}&=\sum_{\underline{h}\in\mathcal{H}\backslash (1,1)}\mu_{\underline{h}}1\{h_{i}=\underline{h}\}+\sum_{\underline{h}\in\mathcal{H}_{S}}\Delta_{\underline{h}}h_{it}1\{h_{i}=\underline{h}\}+\kappa_{(1,1)}h_{it}1\{\underline{h}=(1,1)\}+\epsilon_{it}.\label{eq:GRC}
\end{align}

where $y$ is log yields,

Table \ref{tbl:Trajectories} maps out the frequencies of the different trajectories present in the data, sorted by frequency. The largest portion of the data are those that do not adopt improved maize seed at all, followed with sharp decline, to those that use improved maize seed in every period. Interestingly, adoption trajectories seem to be more popular for later adoptions than earlier ones.

This method seeks to estimate the following equation:

where y is yields or profits for household i at time t, h is an indicator for adoption,  is a household’s absolute advantage,  is a household’s comparative advantage and  describes the gains to adopting, relative to a household’s comparative advantage. This framework provides an advantage over other methods \citep{Wooldridge1997-xj,Heckman1998-pt}. Firstly, identification does not rely on the existence of a valid instrumental variable. Instead, identification comes from a linear projection of an individual's returns to adoption onto the observed history of their adoption, an approach similar to the correlated random effects (CRE) method in \citep{Chamberlain1984-uk}. Secondly, it disentangles a household’s absolute advantage when taking part in agriculture and its comparative advantage in adopting. This is crucial in understanding the adoption decision as it is not dependent on a household’s agricultural ability, but on the gains from taking it up. We and co-authors developed this further in \citep{Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc} by proposing a more flexible and tractable approach using a group random coefficient strategy that draws on recent developments in the nonparametric panel identification literature. Our strategy has a clear economic interpretation and can elucidate potential identification concerns of the CRC model to the practitioner. It is also more easily extended to multiple time periods and new levels of heterogeneity than the CRC approach.

Estimating equation 1 will allow us  to analyze the distribution of the gains to adoption and to create a distribution of heterogeneous returns. We will investigate whether comparative advantage is driven by two key policy relevant factors: water conservation strategies and deviation from historical climate averages. The first will elucidate if adaptation strategies in water management are associated with higher levels of gains to adoption. The second will seek to understand whether households are adapting to climate change through their adoption decisions. This will require the extension of the method developed in \citep{Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc}, by allowing for time-varying characteristics in the identification of comparative advantage.

\subsection{Endogenous Switching regression}
The ESS4 is a unique data-set that includes DNA fingerprinting of maize seed type. This brings new possibilities in terms of exploring.... sayESS4 no longer part of panel but that still useful,....//
The group random coefficient model used the variation in the adoption trajectories of each observation in the ESS 1-3 panel as a source of information that allows for an unbiased estimation. ESS4 is no longer part of this panel data-set, hence, in order to estimate the effect of the adoption of improved seeds on yields, an alternative estimation method is needed. We propose an endogenous switching regression model which accounts for unobserved heterogeneity in adoption decisions. This method, although it imposes restrictive assumptions regarding the distribution of errors

>>>>

%%% Why it is important
Food security is a major challenge in many East African countries, and Ethiopia is no exception, having historically struggled to provide an adequate and reliable food supply \citep{Ramakrishna2002-hv, Jaleta2018-oj}. Since the drought of 1984, Ethiopia has become increasingly reliant on maize, as of today, it constitutes the most widespread crop in the country.\footnote{FAO stats data shows the total area cultivated with Maize in 2019 is 25\% larger than the one cultivated with the next most widespread crop (sorghum), and 27\% and 139\% larger than the next two following crops, wheat and Barley, respectively.} As Figure \ref{fig:maize_yields} shows, there is an upward trend in the total area cultivated with maize since 1990, with average yields that have almost doubled during this period.

%%%%% The puzzle 
Despite numerous technological breakthroughs in maize germplasm with increased yields and drought tolerance, adoption of improved seeds has remained a challenge, with only 13\% of maize farmers consistently using improved seeds varieties, and the lion share of farmers (62\%) not using improved seeds at all during the period where data is available. Moreover, there is a large percentage of farmers who adopt new varieties inconsistently limiting their opportunities to learn about their benefits, see Table \ref{tbl:Trajectories}.

@@

The advantage of our estimation strategy is that we do not rely on identification from a valid instrument, but modelling the heterogeneity directly through a projection of their returns to adoption on their observed "trajectory" of adoption. This allows us to directly capture two effects: the heterogeneous effects of the so-called "comparative advantage" of adoption, and, under the assumption of linearity in comparative advantage, a description of the overall economy's sorting of farmers based on their comparative advantage. The first is important because it is able to disentangle the effect of a general household's agricultural ability (which can be thought of as a fixed effect) and their adoption-specific potential. The sorting parameter can tell us whether the economy sorts such that those that would benefit the most adoption actually adopt, or whether there are barriers that stop that sorting from happening. See \cite{Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc} for more information on the assumptions involved for the Group Random Coefficients estimator. In our particular case, we contribute to the literature and methodology of \cite{Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc} by providing reasonable ways to aggregate the results of the GRC model, which may seem difficult to interpret in its raw form. We use these aggregations to better understand how adoption affects households across time, as well as whether the frequency of adoption has a strong effect.

Table \ref{tbl:Trajectories} maps out the frequencies of the different trajectories present in the data, sorted by frequency. The largest portion of the data are those that do not adopt improved maize seed at all, followed with sharp decline, to those that use improved maize seed in every period. Interestingly, adoption trajectories seem to be more popular for later adoptions than earlier ones. Most households do not adopt at all (``never-adopters''), but the second most frequent trajectory are those that adopt in every wave (``always-adopters''). Let $\mathcal{H}$ be the set of these trajectories in the data.

@@

\begin{align}
y_{it}&=\sum_{\underline{h}\in\mathcal{H}\backslash (1,1,1)}\mu_{\underline{h}}+\Delta_{(0,0,1)}h_{it}+\phi(\mu_{(0,0,1)}-\mu_{(0,0,1)})h_{it}1\{h_{i}=(0,0,1)\}\nonumber\\
&~~~+\left(\mu_{(1,1,1)}+\phi\left(\mu_{(1,1,1)}-\mu_{(0,0,1)}\right)\right)h_{it}1\{h_{i}=(1,1,1)\} + X_{it}\beta +\varepsilon_{it},\label{eq:GRC_Suri}
\end{align}

where $y$ is log yields, $\mu_{\underline{h}}$ are binary variables for being in trajectory $\underline{h}$, $h$ is a binary variable for whether household $i$ adopted in wave $t$, $X$ are a set of controls and $\Delta_{0,0,1}$ are the returns to adoption for trajectory $(0,0,1)$. For reasons outlined in \cite{Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc}, we use a base trajectory. $\phi$ is the parameter which summarizes sorting into the economy. If $\phi$ is positive, then higher yields is associated with higher comparative advantage, denoting that those that adopt also tend to have the highest returns to adoption. If it is negative, however, this denotes that there might be barriers, such as a lack of fertilizer access, and bad roads that restrict the ability to buy hybrid seed.

Equation \ref{eq:GRC_Suri} can be estimated by way of General Method of Moments. Parameters for returns to adoption and comparative advantage can be post-estimated by the delta method, where the expression for returns of each trajectory can be derived from:

$$
\Delta_{\underline{h}}=\phi\left(\mu_{\underline{h}}-\mu_{(0,0,1)}\right) + \Delta_{(0,0,1)}
$$

And comparative advantage can be derived from:

$$
\theta_{\underline{h}} = \mu_{\underline{h}} - E(\mu_{\underline{h}})
$$

where $E(\mu_{\underline{h}}) = \sum_{\mathcal{H}} \pi_h \cdot \mu_h$ for $\pi_h$ being the frequency weight for trajectory $h$. The comparative advantage of each trajectory tells us how much a household with a given trajectory benefits from adoption. $\theta <0 $ denotes that returns to adoption are negative and that a household loses by adopting.

Once we generate $\theta$ and $\Delta$ for each trajectory, we can examine heterogeneous effects to adoption. In its raw form

\subsection{Endogenous Switching regression}
The ESS4 is a unique data-set that includes DNA fingerprinting of maize seed type. This brings new possibilities in terms of exploring.... sayESS4 no longer part of panel but that still useful,....//
The group random coefficient model used the variation in the adoption trajectories of each observation in the ESS 1-3 panel as a source of information that allows for an unbiased estimation. ESS4 is no longer part of this panel data-set, hence, in order to estimate the effect of the adoption of improved seeds on yields, an alternative estimation method is needed. We propose an endogenous switching regression model which accounts for unobserved heterogeneity in adoption decisions. This method, although it imposes restrictive assumptions regarding the distribution of the errors
